Inferring the gravitational potential of the Milky Way with a few precisely measured stars

### 2014-02-12 text/paper2/ms.tex
kvj structure

- Broader comparison to other methods\\
\indent\indent- 'generative methods' more generally (ie N-body, plus Varghese, plus Kuepper plus new possible action-angle models....\\
\indent\indent- statistical methods\\
\indent\indent- Sanders \& Binney 2013\\
- Then a short note on worries\\
\indent\indent- for generative models\\
\indent\indent- for any methods based on action/angle \\
\indent\indent- GD-1 was fit as if it highlighted an orbit.  That is known to be wrong.\\
\indent\indent- Densities are hard to match: need a background model, way to go from luminosity to density, needs assumptions about the mass function and structure of the satellite\\
\indent\indent- Triaxiality / radially-dependent shapes and substructure present problems for integral-based methods\\
- Then how rewinder gets round these????\\
\indent\indent- Reference Deg \& Widrow to say they illustrate the idea of the power of multiple streams to back up the bit where you say that's the long-term aim\\

@@

[Chemical tagging will always only give probabilistic information.
  But what if it delivers somewhat confident information about small
  stellar families in the halo?]

@@

\section{The model}

@@

[Conceptual description of the model] Figure~\ref{fig:reldist}

Take $\D$ to be the observed, heliocentric 6D position of a star --- e.g., the measured position on the sky, ($l$, $b$); distance, $D$; proper motions, ($\mu_l$, $\mu_b$); and line-of-sight velocity, $v_r$ --- and take $\W$ to be the corresponding \emph{true}, 6D position of the star in the same coordinate and reference frame. Assume that we have determined that this star was once part of a progenitor system (e.g., satellite galaxy) tidally disrupting and forming a cold debris structure in the potential, $\Phi$, of some parent galaxy. The position of the progenitor is observed to be at heliocentric 6D position $\D_p$, with true position $\W_p$ (where any subscript $p$ refers to the progenitor). The true 6D heliocentric spherical positions of the star and progenitor -- $\W$ and $\W_p$ -- have corresponding 6D Galactocentric cartesian coordinate vectors $\X$ and $\X_p$ comprised of 3D position and velocity vectors ($\bsr$,$\bsv$) and ($\bsr_p$,$\bsv_p$). Initially we assume the present-day mass of the progenitor, $M_p$, and velocity dispersion, $\sigma_v$, have been measured perfectly and neglect mass-loss from the progenitor. With this notation, we write the joint posterior probability for the observed position of a star, progenitor, and parent potential parameters as:
\begin{align}
	p(\Phi, \W, \W_p, \tau \given \D, \D_p) &= \frac{1}{\mathcal{Z}} p(\D \given \W) p(\D_p \given \W_p) 
												       p(\W \given \W_p, \tau, \Phi) 
												       p(\Phi) p(\tau)
\end{align}
where the factor $\mathcal{Z}$ only depends on the properties of the data (the \emph{evidence}), and is thus constant when varying the model parameters. We assume that the observational errors are Gaussian in heliocentric coordinates:
\begin{align}
	p(\D \given \W) &= \mathcal{N}(\W \given \D, \bSigma)\\
	p(\D_p \given \W_p) &= \mathcal{N}(\W_p \given \D_p, \bSigma_p)
\end{align}
where the covariance matrices $\bSigma$ and $\bSigma_p$ specify the observational uncertainties on the observed 6D position of star and the progenitor, respectively.

@@

The likelihood of a star's position today is computed by first transforming the observed position to Galactocentric cartesian coordinates and integrating the orbits of the progenitor and star backwards for several Gigayears in the parent potential, $\Phi$, treating both as test particles. The distance, $R$, and velocity, $V$, of the star relative to the progenitor are computed at each time-step, and the likelihood is computed as the product of the isotropic, log-normal centered on the logarithm of the instantaneous tidal radius with the isotropic, log-normal centered on the logarithm of the velocity dispersion of the progenitor:
\begin{align}
	p(\W \given \W_p, \tau, \Phi) &= p(\X \given \X_p, \tau, \Phi) \jac{(l,b,D,\mu_l,\mu_b,v_r)}_{t=\tau}\\
	p(\X \given \X_p, \tau, \Phi) &= p(\ln R, \ln V \given \tau, \Phi) \jac{(x,y,z,v_x,v_y,v_z)}_{t=\tau}\label{eq:int} \\
	p(\ln R, \ln V \given \tau, \Phi) &= \left[\mathcal{N}(\ln R | \ln \rtide, \sigma_R)\mathcal{N}(\ln V | \ln \sigma_v, \sigma_V)\right]_{t=\tau}
\end{align}
% , but the vectors $\D$ and $\W$ [exist] in different coordinate systems --- heliocentric spherical and Galactocentric cartesian, respectively --- so we must also track the Jacobian that [describes] this transformation:
where in eq.~\ref{eq:int} we have implicitly integrated over and neglected the two angle coordinates associated with $\ln R$ and the two angles associated with $\ln V$. 
\begin{align}
	\ln p(\ln R, \ln V \given \tau, \Phi) = 
\end{align}
We then compute the marginal likelihood, $p(\W \given \W_p, \Phi)$, by integrating over all possible unbinding times,
\begin{align}
	p(\W \given \W_p, \Phi) = \int p(\W \given \W_p, \tau, \Phi)p(\tau) d\tau.
\end{align}
In general, $p(\tau)$ is likely conditional on the orbit of the progenitor --- stars are preferentially stripped at pericentric passages --- but this is neglected and is taken to be uniform over the entire interaction history between the progenitor and parent.

\section{Experiments}
We conduct several tests of this method using N-body simulations run with an SCF code [...]. [KVJ: describe simulations here?]

>>>>

% Generate approx. to N-body -> single Orbit --> Orbits + offsets --> Bovy
% method, problem, etc. then address what we fix

- Broader comparison to other methods\\
\indent\indent- 'generative methods' more generally (ie N-body, Bovy, plus Varghese, plus Kuepper plus new possible action-angle models....\\
\indent\indent- Sanders \& Binney 2013\\
\indent\indent- statistical methods, Penarrubia, Sanderson, Rewinder\\
- Then a short note on worries\\
\indent\indent- for generative models\\
\indent\indent- for any methods based on action/angle \\
\indent\indent- GD-1 was fit as if it highlighted an orbit.  That is known to be wrong.\\
\indent\indent- Densities are hard to match: need a background model, way to go from luminosity to density, needs assumptions about the mass function and structure of the satellite\\
\indent\indent- Triaxiality / radially-dependent shapes and substructure present problems for integral-based methods\\
- Then how rewinder gets round these????\\
\indent\indent- Reference Deg \& Widrow to say they illustrate the idea of the power of multiple streams to back up the bit where you say that's the long-term aim\\

@@

%[Chemical tagging will always only give probabilistic information.
%  But what if it delivers somewhat confident information about small
%  stellar families in the halo?]

@@

\section{Simulations}\label{sec:sims}

We [...] N-body simulations run with an SCF code [...]. [KVJ: describe simulations here?]

\section{...}

\subsection{Notation}
Take $\D$ to be the observed, heliocentric 6D position of a star --- e.g., the measured position on the sky, ($l$, $b$); distance, $D$; proper motions, ($\mu_l$, $\mu_b$); and line-of-sight velocity, $v_r$ --- and take $\W$ to be the corresponding \emph{true}, 6D position of the star in the same coordinate and reference frame. Assume that we have determined that this star was once part of a progenitor system (e.g., satellite galaxy) tidally disrupting and forming a cold debris structure in the potential, $\Phi$, of some parent galaxy. The position of the progenitor is observed to be at heliocentric 6D position $\D_p$, with true position $\W_p$ (where any subscript $p$ refers to the progenitor). The true 6D heliocentric spherical positions of the star and progenitor -- $\W$ and $\W_p$ -- have corresponding 6D Galactocentric cartesian coordinate vectors $\X$ and $\X_p$ comprised of 3D position and velocity vectors ($\bsr$,$\bsv$) and ($\bsr_p$,$\bsv_p$). Initially we assume the present-day mass of the progenitor, $M_p$, and velocity dispersion, $\sigma_v$, have been measured perfectly and neglect mass-loss from the progenitor.

\subsection{Motivation}
In \cite{apw13}, we introduced the \emph{Rewinder} algorithm [...]. In that work, we define a phase-space distance, $D_{ps}$, as the sum of the normalized, relative position, $\frac{\bsr-\bsr_p}{\rtide}$, and velocity, $\frac{\bsv-\bsv_p}{v_{\rm esc}}$, for a stream star. We minimized the generalized variance of the 6D distribution over each normalized coordinate evaluated at the time that each star has minimum $D_ps$. This simple idea proved quite powerful for constraining properties of the LM10 potential used simulated observations of their N-body simulation. However, this method does not properly incorporate observational uncertainties.

Using the simulations described in Section~\ref{sec:sims}

@@

%The core of this method provides a basis for the probabilistic model presented below.

\subsection{Probabilistic model}
With this notation, we write the joint posterior probability for the observed position of a star, progenitor, and parent potential parameters as:
\begin{align}
	p(\Phi, \W, \W_p, \tau \given \D, \D_p) &= \frac{1}{\mathcal{Z}} p(\D \given \W) p(\D_p \given \W_p) 
												       p(\W \given \W_p, \tau, \Phi) 
												       p(\Phi) p(\tau)
\end{align}
where the factor $\mathcal{Z}$ only depends on the properties of the data (the \emph{evidence}), and is thus constant when varying the model parameters. We assume that the observational errors are Gaussian in heliocentric coordinates:
\begin{align}
	p(\D \given \W) &= \mathcal{N}(\W \given \D, \bSigma)\\
	p(\D_p \given \W_p) &= \mathcal{N}(\W_p \given \D_p, \bSigma_p)
\end{align}
where the covariance matrices $\bSigma$ and $\bSigma_p$ specify the observational uncertainties on the observed 6D position of star and the progenitor, respectively.

@@

The likelihood of a star's position today is computed by first transforming the observed position to Galactocentric cartesian coordinates and integrating the orbits of the progenitor and star backwards for several Gigayears in the parent potential, $\Phi$, treating both as test particles. The distance, $R$, and velocity, $V$, of the star relative to the progenitor are computed at each time-step, and the likelihood is computed as the product of the isotropic, log-normal centered on the logarithm of the instantaneous tidal radius with the isotropic, log-normal centered on the logarithm of the velocity dispersion of the progenitor:
\begin{align}
	p(\W \given \W_p, \tau, \Phi) &= p(\X \given \X_p, \tau, \Phi) \jac{(l,b,D,\mu_l,\mu_b,v_r)}_{t=\tau}\\
	p(\X \given \X_p, \tau, \Phi) &= p(\ln R, \ln V \given \tau, \Phi) \jac{(x,y,z,v_x,v_y,v_z)}_{t=\tau}\label{eq:int} \\
	p(\ln R, \ln V \given \tau, \Phi) &= \left[\mathcal{N}(\ln R | \ln \rtide, \sigma_R)\mathcal{N}(\ln V | \ln \sigma_v, \sigma_V)\right]_{t=\tau}
\end{align}
% , but the vectors $\D$ and $\W$ [exist] in different coordinate systems --- heliocentric spherical and Galactocentric cartesian, respectively --- so we must also track the Jacobian that [describes] this transformation:
where in eq.~\ref{eq:int} we have implicitly integrated over and neglected the two angle coordinates associated with $\ln R$ and the two angles associated with $\ln V$. 
\begin{align}
	\ln p(\ln R, \ln V \given \tau, \Phi) = -\ln(2\pi)-\ln(\sigma_R\sigma_V)-
		\ln\left[\left(\frac{R}{\rtide}\right)^{1/\sigma_R} \left(\frac{V}{\sigma_v}\right)^{1/\sigma_V}\right]
\end{align}
We then compute the marginal likelihood, $p(\W \given \W_p, \Phi)$, by integrating over all possible unbinding times,
\begin{align}
	p(\W \given \W_p, \Phi) = \int p(\W \given \W_p, \tau, \Phi)p(\tau) d\tau.
\end{align}
In general, $p(\tau)$ is likely conditional on the orbit of the progenitor --- stars are preferentially stripped at pericentric passages --- but this is neglected and is taken to be uniform over the entire interaction history between the progenitor and parent.

\section{Experiments}
